DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life

### 2022-02-18 manuscript/manuscript.Rmd
working on figure 2

\item `datelife` currently accepts scientific names only as input, and they can belong any taxonomic group or binomial specific. If an input taxon name belongs to a "higher" taxonomic group, i.e., a taxon above the species level, such as genus, family, etc.), `datelife` has two alternative behaviors defined by the "get species from taxon" flag. If the flag is active, `datelife` retrieves all species names within the "higher" taxonomic group and adds them to the input. If the flag is inactive, `datelife` will drop the "higher" taxonomic names from input.
\item The cleaned input taxon names are saved as a `datelifeQuery` object
and used to search the chronogram database.
\item Chronograms with at least two matching input taxon names are identified and pruned down to preserve only input taxon names as tips.
Then, each pruned chronogram is transformed to a patristic distance matrix. This format facilitates and greatly speeds up all downstream analyses and summaries. The matrices are associated to the citation of the original study and stored as a `datelifeResult` object.
\item  At this point, various summary data can be obtained to inform decisions for the next steps of the analysis workflow. Types of summary information provided are: a) all pruned source chronograms, b) age of the MRCA (most recent common ancestor) of the pruned source chronograms, c) citations of studies where pruned source chronograms were originally published, d) a summary table with all of the above, e) a single summary chronogram of all or a subset of pruned source chronograms, f) a report of successful matches of input taxon names across pruned source chronograms, and g) the single pruned source chronogram with the most matching input taxon names.
\item  Finally, time of lineage divergence obtained from the pruned chronograms can be used as secondary calibration points to date a tree with or without branch lengths containing some or all input taxon names. %<!--, a taxonomic tree-->
\item  If there is no information available for any input taxon name, users can also create both age and phylogenetic data for the missing branches with a variety of algorithms described below.
\item  Users can easily save all source and summary chronograms in formats that permit easy reuse and reanalyses (newick and R "phylo" format), as well as view and compare results graphically, or construct their own graphs using \texttt{datelife}'s graphic generation functions.
\end{enumerate}

>>>>

\item `datelife` currently accepts scientific names only as input, and they can belong any taxonomic group or binomial specific. If an input taxon name belongs to a "higher" taxonomic group, i.e., a taxon above the species level, such as genus, family, etc.), `datelife` has two alternative behaviors defined by the "get species from taxon" flag. If the flag is active, `datelife` retrieves all species names within the "higher" taxonomic group and adds them to the input. If the flag is inactive, `datelife` will drop the "higher" taxonomic names from input.
\item The cleaned input taxon names are saved as a `datelifeQuery` object
and used to search the chronogram database.
\item Chronograms with at least two matching input taxon names are identified and pruned down to preserve only input taxon names as tips.
Then, each pruned chronogram is transformed to a patristic distance matrix. This format facilitates and greatly speeds up all downstream analyses and summaries. The matrices are associated to the citation of the original study and stored as a `datelifeResult` object.
\item  At this point, various summary data can be obtained to inform decisions for the next steps of the analysis workflow. Types of summary information provided are: a) all pruned source chronograms, b) age of the MRCA (most recent common ancestor) of the pruned source chronograms, c) citations of studies where pruned source chronograms were originally published, d) a summary table with all of the above, e) a single summary chronogram of all or a subset of pruned source chronograms, f) a report of successful matches of input taxon names across pruned source chronograms, and g) the single pruned source chronogram with the most matching input taxon names.
\item To construct summary trees from patristic distance matrices, one can use a clustering method like Neighbor Joining. We implemented this in the package. However, it is know that these methods can return toologies that are not biologically plausible (Example and citation). Hence, we implement
a different workflow, in which we use a fixed topology taken from the literature or from expert
phylogenetic information (such as the OpenTree synthetic tree), and we use the patristic distances as calibrations to date that topology (with BLADJ for now). TODO ADD FIGURE: mock example explaining differenc ebetween using summary ages and raw ages to calibrate a topology. Then show it in the biological example. This will show what happens if a tree (phylogenetic conflict) or node (age) do not agree.

\item  Alternatively, time of lineage divergence obtained from the pruned chronograms can be used directly as secondary calibration points to date a tree with or without branch lengths containing some or all input taxon names. %<!--, a taxonomic tree-->
\item  If there is no information available for any input taxon name, users can also create both age and phylogenetic data for the missing branches with a variety of algorithms described below.
\item  Users can easily save all source and summary chronograms in formats that permit easy reuse and reanalyses (newick and R "phylo" format), as well as view and compare results graphically, or construct their own graphs using \texttt{datelife}'s graphic generation functions.
\end{enumerate}
